{\scshape Denicek}: Computational Substrate for\\ Document-Oriented End-User Programming

### 2025-04-04 paper.tex
Spellcheck

A computational substrate defines the structures with which programs are constructed, how
the progam state is represented and how the state evolves during execution \cite{jakubovic-2022-ladder}.
The choice of a substrate affects what programming experiences can be readily supported.
For example, object-oriented programming has been historically linked to graphical user
interfaces~\cite{kay-1993-smalltalk}, while representing programs as lists enabled Lisp to become
a language laboratory \cite{steele-1993-lisp}.

@@

\begin{itemize}
\item \emph{Collaborative editing.} Users should be able to locally modify a shared document
  and merge their changes, preferrably in a way that does not require a live connection to a central server~\cite{kleppmann-2019-local}.
\item \emph{Programming by Demonstration.} Allow users to construct simple programs by
  enacting the steps of the desired behavior using concrete examples and generalizing from the example~\cite{leiva-2021-rapido,cypher-1993-pbd}.
\item \emph{Incremental Recomputation.} When a part of a document changes, formulas whose result
  depends on the part are invalidated and, possibly automatically, recomputed \cite{teitelbaum-1981-cps,mcdirmid-2013-usable,horowitz-2023-engraft}.
\item \emph{Schema Change Control.} When the user evolves the structure of the document, affected
  data and formulas should automatically co-evolve to match the new structure~\cite{litt-2020-cambria,edwards-2025-schema}.
\item \emph{End-User Debugging.} The user should be able to ask provenance questions~\cite{cheney-2009-provenance} to understand
  why a computation resulted in a particular value and what inputs contributed to the result \cite{ko-2009-whyline}.
\item \emph{Concrete Programming.} It should be possible to reuse parts of program logic, or formulas,
  without introducing abstractions, that is, program against concrete values \cite{edwards-2006-copypaste,edwards-2022-copypaste}.
\end{itemize}

@@

\begin{itemize}
\item We present the Denicek substrate (\S\ref{sec:system}) and provide a detailed description of its
  document representation, edit operations and the key three operations for working with edit histories.
\item We illustrate a range of end-user programming experiences supported in Webnicek, a simple
  web-based prototype programming system (\S\ref{sec:walk}), and discuss how the experiences are
  implemented on top of the Denicek substrate (\S\ref{sec:impl}).
\item We document important design decisions, alternatives and limitations (\S\ref{sec:discuss}).
  The analysis shows that the desired functionality requires a careful choice among interconnected
  design options.
\item To evaluate how Denicek simplifies the development of program\-ming systems, we
  build an innovative data science notebook Datnicek and assess its impementation
  complexity~(\S\ref{sec:case}). We also present heuristic evaluation of the system (\S\ref{sec:eval}).
\end{itemize}

@@

\subsubsection*{Edit Histories and Merging}
Manipulating programs through semantically meaningful edits is a technique used by structure
editors \cite{teitelbaum-1981-cps,hempel-2018-deuce,beckman-2023-sandblocks}. The langauge of
edits has been captured formally as an \emph{edit calculus} \cite{omar-2017-hazelnut} and
edit histories have also been recognized as a suitable basis for live programming environments \cite{storm-2013-deltas}.

Merging of edits is most frequently done in version control systems such as git. Source control ystems
based on a more rigorous design such as Pijul \cite{meunier-2024-pijul} delay merging to a later point
by using a graph or a lattice \cite{schurmann-2022-merging}. In the context of
programming environments, Grove \cite{adams-2025-grove} uses a commutative patches, with a graph
structure similar to Pijul, as the basis for a collaborative structure editor.

@@

\subsubsection*{Concrete Programming}
Programming can be simplified by working with concrete values instead of abstractions, an idea
pioneered by the prototype-based programming language Self \cite{ungar-1987-self}. In Self,
prototypes are used at the object level. At the expression level, similar functionality can be provided
by managed copy \& paste. Subtext \cite{edwards-2006-copypaste,edwards-2022-copypaste} treats this
mechanism as central, whereas other systems view tracking of copy \& paste as an extra
editor featrue \cite{jablonski-2007-cren,toomim-2004-linked}. Copy \& paste has also been
tracked in spreadsheets \cite{hermans-2015-copypaste}. Gridlets \cite{joharizadeh-2020-gridlets}
offer a spreadsheet abstraction based on reusing concrete computation, similar to the one proposed
for The Gamma~\cite{petricek-2020-live}.

\subsubsection*{Other Programming Experiences}
Several experiences not directly addressed in this paper are worth further investigation.
Projectional editors such as Lorgnette \cite{gobert-2023-lorgnette},
live literals \cite{omar-2021-livelits}, projection boxes \cite{lerner-2020-boxes} and data detectors
\cite{nardi-1998-agents} allow visualization or editing of aspects of programs through a user interface.
The problem of making live and rich editors compositional is addressed by
Engraft \cite{horowitz-2023-engraft}. Programming environments may be also improved by integrating
live examples \cite{rauch-2019-babylonian} and a variety of AI assistantce
tools \cite{petricek-2023-aias,blinn-2024-llms,mcnutt-2023-nbai}.

@@

\begin{tabular}{|cl|}
\hline
\rowcolor{ekgray}
&\\[-1em]
\rowcolor{ekgray}
& \sffamily\small{\bfseries Kind}\;\,\textit{\footnotesize arguments} \\[0.2em]
\hline
&\\[-1em]
\ndtablecol{\faListUl}{List}{tag, index$_1$, child$_1$, $\ldots$, index$_n$, child$_n$}
  {Ordered list of nodes, addressable by \textit{index}. Renders as \Verb|<tag>| with children.}
\ndtablecol{\faFileO}{Record}{tag, field$_1$, child$_1$, $\ldots$, field$_n$, child$_n$}
  {Record with children addressable by \textit{field}. Renders as \Verb|<tag>| with children.}
\ndtablecol{\faExternalLink}{Reference}{selectors}
  {Reference to another document location. Displays the \Verb|selectors| as a link.}
\ndtablecol{\faFont}{Primitivie}{string \textnormal{or} number}
  {Numerical or textual primitive value. Renders as an HTML text node.}
\hline
\end{tabular}
\vspace{-0.5em}
\caption{Structure of selectors and document nodes}
\label{fig:doc}
\vspace{-1em}
\end{figure}

@@

Another use of value edits is when evaluating a formula, which involves wrapping the formula
node (\S\ref{sec:impl-eval}), an operation that should not affect references to the formula
itself. To support value edits, edits that normally transform references (Fig.~\ref{fig:edits} below)
have a \emph{reference behavior} option that can disable automatic reference
updating. This annotation is required when the target contains the \emph{Index} selector,
which is necessarilly a value edit affecting a specific node.

@@

\subsection{Collaboratative Editing}
\label{sec:impl-collab}

@@

Second, when replaying edits, Webnicek does not append the recorded edits on top of the current
history. Instead, it stores the hash of the history at the time of saving. To replay the edits,
it then appends the edits to the top of the original history and merges this edit sequence
with the current history. This pushes the recorded edits through later edits made by the user,
reconciling them with potential structural edits. The result can be seen in Fig.~\ref{fig:walkthrough} (E), where
a newly added list item becames a table row.

@@

Denicek also only supports the \emph{convergence} model of collaborative editing \cite{edwards-2025-schema}
where users have to merge all changes in order and cannot adopt selected edits (cherry
picking in git). The \emph{divergence} model would let users keep their own schema but
import all data edits. Supporting the model requires a \emph{retract} operation \cite{edwards-2021-typed}
that is dual to our edit reconcilliation (given subsequent edits $e_1, e_2$, generate
$e_2'$ that has the same effect as $e_2$ but can occur before $e_1$).

@@

\subsubsection*{Effectiveness and Unifromity.}
Many of the Datnicek requirements are addressed by Denicek with minimal development
effort. Denicek also provides a small and uniform set of primitives. In particular,
the implementation represents all non-transient changes to a notebook as Denicek edits.
Code completions offered by the code editor are edits (wrapping a formula or appending a parameter),
changing a primitive value is an edit, and recommendations in the interactive grid are also edits
(appending a {\small\Verb_<transform>_} node).

@@

Our experience with conflict detection shows the importance of getting the details of the
implementation of programming susbstrates like Denicek right. This could be aided by the
development of a formally tractable model of the system and proving a correctness property
akin to that of The Gamma live previews \cite{petricek-2020-live}.

>>>>

A computational substrate defines the structures with which programs are constructed, how
the program state is represented and how the state evolves during execution \cite{jakubovic-2022-ladder}.
The choice of a substrate affects what programming experiences can be readily supported.
For example, object-oriented programming has been historically linked to graphical user
interfaces~\cite{kay-1993-smalltalk}, while representing programs as lists enabled Lisp to become
a language laboratory \cite{steele-1993-lisp}.

@@

\begin{itemize}
\item \emph{Collaborative editing.} Users should be able to locally modify a shared document
  and merge their changes, preferably in a way that does not require a live connection to a central server~\cite{kleppmann-2019-local}.
\item \emph{Programming by Demonstration.} Allow users to construct simple programs by
  enacting the steps of the desired behavior using concrete examples and generalizing from the example~\cite{leiva-2021-rapido,cypher-1993-pbd}.
\item \emph{Incremental Recomputation.} When a part of a document changes, formulas whose result
  depends on the part are invalidated and, possibly automatically, recomputed \cite{teitelbaum-1981-cps,mcdirmid-2013-usable,horowitz-2023-engraft}.
\item \emph{Schema Change Control.} When the user evolves the structure of the document, affected
  data and formulas should automatically co-evolve to match the new structure~\cite{litt-2020-cambria,edwards-2025-schema}.
\item \emph{End-User Debugging.} The user should be able to ask provenance questions~\cite{cheney-2009-provenance} to understand
  why a computation resulted in a particular value and what inputs contributed to the result \cite{ko-2009-whyline}.
\item \emph{Concrete Programming.} It should be possible to reuse parts of program logic, or formulas,
  without introducing abstractions, that is, program against concrete values \cite{edwards-2006-copypaste,edwards-2022-copypaste}.
\end{itemize}

@@

\begin{itemize}
\item We present the Denicek substrate (\S\ref{sec:system}) and provide a detailed description of its
  document representation, edit operations and the key three operations for working with edit histories.
\item We illustrate a range of end-user programming experiences supported in Webnicek, a simple
  web-based prototype programming system (\S\ref{sec:walk}), and discuss how the experiences are
  implemented on top of the Denicek substrate (\S\ref{sec:impl}).
\item We document important design decisions, alternatives and limitations (\S\ref{sec:discuss}).
  The analysis shows that the desired functionality requires a careful choice among interconnected
  design options.
\item To evaluate how Denicek simplifies the development of program\-ming systems, we
  build an innovative data science notebook Datnicek and assess its implementation
  complexity~(\S\ref{sec:case}). We also present heuristic evaluation of the system (\S\ref{sec:eval}).
\end{itemize}

@@

\subsubsection*{Edit Histories and Merging}
Manipulating programs through semantically meaningful edits is a technique used by structure
editors \cite{teitelbaum-1981-cps,hempel-2018-deuce,beckman-2023-sandblocks}. The language of
edits has been captured formally as an \emph{edit calculus} \cite{omar-2017-hazelnut} and
edit histories have also been recognized as a suitable basis for live programming environments \cite{storm-2013-deltas}.

Merging of edits is most frequently done in version control systems such as git. Source control systems
based on a more rigorous design such as Pijul \cite{meunier-2024-pijul} delay merging to a later point
by using a graph or a lattice \cite{schurmann-2022-merging}. In the context of
programming environments, Grove \cite{adams-2025-grove} uses a commutative patches, with a graph
structure similar to Pijul, as the basis for a collaborative structure editor.

@@

\subsubsection*{Concrete Programming}
Programming can be simplified by working with concrete values instead of abstractions, an idea
pioneered by the prototype-based programming language Self \cite{ungar-1987-self}. In Self,
prototypes are used at the object level. At the expression level, similar functionality can be provided
by managed copy \& paste. Subtext \cite{edwards-2006-copypaste,edwards-2022-copypaste} treats this
mechanism as central, whereas other systems view tracking of copy \& paste as an extra
editor feature \cite{jablonski-2007-cren,toomim-2004-linked}. Copy \& paste has also been
tracked in spreadsheets \cite{hermans-2015-copypaste}. Gridlets \cite{joharizadeh-2020-gridlets}
offer a spreadsheet abstraction based on reusing concrete computation, similar to the one proposed
for The Gamma~\cite{petricek-2020-live}.

\subsubsection*{Other Programming Experiences}
Several experiences not directly addressed in this paper are worth further investigation.
Projectional editors such as Lorgnette \cite{gobert-2023-lorgnette},
live literals \cite{omar-2021-livelits}, projection boxes \cite{lerner-2020-boxes} and data detectors
\cite{nardi-1998-agents} allow visualization or editing of aspects of programs through a user interface.
The problem of making live and rich editors compositional is addressed by
Engraft \cite{horowitz-2023-engraft}. Programming environments may be also improved by integrating
live examples \cite{rauch-2019-babylonian} and a variety of AI assistance
tools \cite{petricek-2023-aias,blinn-2024-llms,mcnutt-2023-nbai}.

@@

\begin{tabular}{|cl|}
\hline
\rowcolor{ekgray}
&\\[-1em]
\rowcolor{ekgray}
& \sffamily\small{\bfseries Kind}\;\,\textit{\footnotesize arguments} \\[0.2em]
\hline
&\\[-1em]
\ndtablecol{\faListUl}{List}{tag, index$_1$, child$_1$, $\ldots$, index$_n$, child$_n$}
  {Ordered list of nodes, addressable by \textit{index}. Renders as \Verb|<tag>| with children.}
\ndtablecol{\faFileO}{Record}{tag, field$_1$, child$_1$, $\ldots$, field$_n$, child$_n$}
  {Record with children addressable by \textit{field}. Renders as \Verb|<tag>| with children.}
\ndtablecol{\faExternalLink}{Reference}{selectors}
  {Reference to another document location. Displays the \Verb|selectors| as a link.}
\ndtablecol{\faFont}{Primitive}{string \textnormal{or} number}
  {Numerical or textual primitive value. Renders as an HTML text node.}
\hline
\end{tabular}
\vspace{-0.5em}
\caption{Structure of selectors and document nodes}
\label{fig:doc}
\vspace{-1em}
\end{figure}

@@

Another use of value edits is when evaluating a formula, which involves wrapping the formula
node (\S\ref{sec:impl-eval}), an operation that should not affect references to the formula
itself. To support value edits, edits that normally transform references (Fig.~\ref{fig:edits} below)
have a \emph{reference behavior} option that can disable automatic reference
updating. This annotation is required when the target contains the \emph{Index} selector,
which is necessarily a value edit affecting a specific node.

@@

\subsection{Collaborative Editing}
\label{sec:impl-collab}

@@

Second, when replaying edits, Webnicek does not append the recorded edits on top of the current
history. Instead, it stores the hash of the history at the time of saving. To replay the edits,
it then appends the edits to the top of the original history and merges this edit sequence
with the current history. This pushes the recorded edits through later edits made by the user,
reconciling them with potential structural edits. The result can be seen in Fig.~\ref{fig:walkthrough} (E), where
a newly added list item becomes a table row.

@@

Denicek also only supports the \emph{convergence} model of collaborative editing \cite{edwards-2025-schema}
where users have to merge all changes in order and cannot adopt selected edits (cherry
picking in git). The \emph{divergence} model would let users keep their own schema but
import all data edits. Supporting the model requires a \emph{retract} operation \cite{edwards-2021-typed}
that is dual to our edit reconciliation (given subsequent edits $e_1, e_2$, generate
$e_2'$ that has the same effect as $e_2$ but can occur before $e_1$).

@@

\subsubsection*{Effectiveness and Uniformity.}
Many of the Datnicek requirements are addressed by Denicek with minimal development
effort. Denicek also provides a small and uniform set of primitives. In particular,
the implementation represents all non-transient changes to a notebook as Denicek edits.
Code completions offered by the code editor are edits (wrapping a formula or appending a parameter),
changing a primitive value is an edit, and recommendations in the interactive grid are also edits
(appending a {\small\Verb_<transform>_} node).

@@

Our experience with conflict detection shows the importance of getting the details of the
implementation of programming substrates like Denicek right. This could be aided by the
development of a formally tractable model of the system and proving a correctness property
akin to that of The Gamma live previews \cite{petricek-2020-live}.
